### 2022-11-10 NSF2022/Description/specificaim2.tex
changes to  background

@@

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}.

When granules are close to one another, which is inevitable in dense and
confined suspensions, the resulting integrals are nearly-singular. This
issue is well-known in the IE community and several algorithms have been
developed in the last 30 years, including panel-based
methods~\cite{helsing_2008a, helsing_tutorial_2012}, Quadrature by
Expansions (QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013,
ste-bar2022, af2018sisc}, Barycentric-type formulae for Cauchy
integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991}, and
interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

In two-dimensions, analytic function theory can be used to derive a
quadrature method whose error is uniformly bounded~\cite{bar2014,
bar-wu-vee2015, ioa-pap-per1991}, but these works only show how to apply
the method to the two-dimensional Laplace and Stokes equations. PI BQ
recently showed that these quadrature methods can be extended to other
PDEs with singularity subtraction~\cite{che-lin-her-qua2022}. For
example, the fundamental solution of the screened Laplace
equation~\eqref{eqn:phase} involves the modified Bessel function of the
second kind, $K_0(\|\xx - \yy\|)$. The behavior of this function as $\yy
\rightarrow \xx$ is $K_0(\alpha \|\xx - \yy\|) = \log\|\xx - \yy\| +
S(\xx - \yy)$, where $S(\xx - \yy)$ is continuously differentiable. This
suggests a quadrature scheme for where the aforementioned quadrature
method is applied to the $\log\|\xx - \yy\|$ kernel, and high-order
methods~\cite{tre-wei2014} are applied to integrals involving the smooth
kernel $S(\xx - \yy)$.Similar techniques, which they call subtraction
techniques, have been considered by Camille Carvalho and her
colleagues~\cite{car2021, car2020, car-kha-kim2018}. These methods are
particularly powerful when computing gradients of layer potentials which
is required, for example, when computing the water-order
stress~\eqref{eqn:stressbalance}.

In three dimensions, the quadrature methods are much more challenging.
Singularity subtraction methods can be applied, but analytic function
theory is no longer applicable. Singularity subtraction methods have
been applied using, for example, smoothed variations of
the~\cite{beale1, cor2001, tlu-bea2019}. Alternatively, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}. However, since we are interested in round (2D) and
spherical (3D) granules, we will exploit the fact that integrals can be
computed exactly after transforming the integrals to to a Fourier (2D)
or spherical harmonic (3D) basis. This technique can be used to evaluate
layer potentials for both Stokes~\cite{cor-vee2018} and the screened
Laplace equation~\cite{koh-cor-che-vee2021}.

Another facet to consider when choosing the quadrature rule is its
ability to be applied efficiently. Since carefully chosen IE
formulations require a mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}, the required CPU time is
proportional to the cost of a matrix-vector multiplication that can be
performed in optimal or near-optimal time with the fast multipole method
(FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3, fmm4, fmm6,
fmm7, ros-ols2016}. PI BQ is experienced with applying FMMs for the
Stokes equation~\cite{qua-bir2014, bys-sha-qua2020} and the screened
Laplace equation~\cite{kro-qua2011, qua2011}. Many of these methods rely
on analytic expansions of the PDE's fundamental solution, but recent
work has demonstrated that a kernel-independent method applied to
suspensions of 80,000 spheres can be simulated with only $\sim$10
seconds per time step on 1792 compute cores~\cite{Yan2019}.

%
%Furthermore, to reduce the
%number of GMRES iterations, the efficacy of preconditioning strategies
%such as the inverse Fast Multipole Method (IFMM)~\cite{cou-pou-dar2017}
%and domain decomposition~\cite{che-bir2021} will be investigated. PI BQ
%used the IFMM to precondition Stokes equations~\cite{qua-cou-dar2018}. A
%suite of other preconditioners including sparse approximate
%inverses~\cite{che2000} and multigrid~\cite{hem-sch1981, sch1982} will
%be investigated.

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}

% ----------------------------------------------------------------------
\subsubsection{Collision-aware suspensions with adaptive time stepping}
\label{subsec:timeStepping}

A numerical issue when simulating the self-assembly of amphiphilic
particles is avoiding particle collision. The hydrophobic attraction
potential drives the amphiphilic particles towards one another and this
leads to physical contact in finite time. Such particle collisions in a
dense, rigid-body suspension are a great challenge and can be a
bottleneck in large-scale simulations. We propose to remedy this
computational challenge by developing novel high-order adaptive time
stepping methods. PI BQ developed a high-order adaptive time
stepping method for hydrodynamic suspensions \cite{qua-bir2016}, and it
has served as a robust method to simulate processes including mixing and
adhesion in suspensions~\cite{qua-vee-you2019, kab-qua-bir2017}.
Following this previous work, the proposed work will use a spectral
deferred correction method~\cite{dut-gre-rok2000} since it iteratively
applies a low-order, single-step method to achieve high-order accuracy.
At each time step, we will compute the total force and torque of the
system which is computationally cheap and physically zero, and therefore
provides an estimate of the local truncation error. PI BQ's experience
is that error estimates based on physical constraints, such as force-
and torque-free conditions, appropriately determine adaptive time step
sizes so that the dynamics are resolved without the computational
expense of embedded Runge-Kutta methods, step-doubling, Richardson
extrapolation.

>>>>

\begin{wrapfigure}[]{l}{2.1in}
  \centering
  \includegraphics[width=2in]{figures/SA2Figures/confinement.pdf}
  \caption{\label{fig:stenosis} A granule-based vesicle passing through
  a constricted geometry.}
\end{wrapfigure}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}. The strong confinement introduces an aditional
computational challenge---acuartely computing integrals between
nearly-touching bodies. This issue is well-known in the IE community and
several algorithms have been developed in the last 30 years, including
panel-based methods~\cite{helsing_2008a, helsing_tutorial_2012},
Quadrature by Expansions (QBX)~\cite{din-hua-mar-tan2021,
klo-bar-gre-one2013, ste-bar2022, af2018sisc}, Barycentric-type formulae
for Cauchy integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991},
and interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

@@

Since we are interested in round (2D) and spherical (3D) granules, we
will exploit the fact that integrals can be computed with spectral
accuracy by transforming the integrals to to a Fourier (2D) or spherical
harmonic (3D) basis. This technique can be used to evaluate layer
potentials for both Stokes~\cite{cor-vee2018} and the screened Laplace
equation~\cite{koh-cor-che-vee2021}. However, for other elliptical and
other shapes, an alternative approach is necessary. The Barycentric-type
formulae for Cauchy integrals results in a uniformly bounded quadrature
error, but the method is described only for the two-dimensional Laplace
and Stokes equations. PI Quaife recently showed that these quadrature
methods can be extended to other two-dimensional
PDEs~\cite{che-lin-her-qua2022}. The method uses singularity subtraction
write the fundamental solution of the PDE as a sum of two terms: a
logarithmic function and a continuously differentiable function. Then
the Barycentric-type formulae can be applied to the logarithmic
function, and the standard trapezoid rule~\cite{tre-wei2014} can be
applied to the other function. These singularity subtraction methods can
also accurately compute derivatives of layer potentials~\cite{car2021,
car2020, car-kha-kim2018} which is required, for example, when computing
the water-order stress~\eqref{eqn:stressbalance}. Quadrature is much
more challenging in three dimensions. Singularity subtraction methods
can be applied, but not combined with Barycentric-type formulae.
Alternatively, smoothed variations of the fundamental solution can be
used to address nearly-singular integrals~\cite{beale1, cor2001,
tlu-bea2019}.  Alternatively, efficient quadrature methods can be
applied directly to accurately compute singular and near-singular
integrals~\cite{khayat_2005, gimbutas2013sisc}.

Another facet to consider when choosing the quadrature rule is its
ability to be applied efficiently. Since carefully chosen IE
formulations require a mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}, the required CPU time is
proportional to the cost of a matrix-vector multiplication that can be
performed in optimal or near-optimal time with the fast multipole method
(FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3, fmm4, fmm6,
fmm7, ros-ols2016}. PI Quaife is experienced with applying FMMs for the
Stokes equation~\cite{qua-bir2014, bys-sha-qua2020} and the screened
Laplace equation~\cite{kro-qua2011, qua2011}. Many of these methods rely
on analytic expansions of the PDE's fundamental solution, but recent
work has demonstrated that a kernel-independent method applied to
suspensions of 80,000 spheres can be simulated with only $\sim$10
seconds per time step on 1792 compute cores~\cite{Yan2019}.

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional reqired
computational tool is a collision-aware adaptive time stepping method.
The order parameter results in the granules attracting one another and
this leads to physical contact in finite time. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions with by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2022}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.
